Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-05 Manuscript/manuscript.Rmd
editted abstract

**Background.** Chromosome conformation capture sequencing technologies have shown that the three-dimensional (3D) structure of the genome is folded into distinct compartments, known as topologically associated domains (TADs) -- units of coordinated gene expression. The location of TAD boundaries has been shown to be highly conserved. Additionally, the distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify the epigenomic features that are most predictive of TAD boundaries.

Current methods rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions that include TAD boundaries. However, many methods ignore key characteristics of domain data, such as the impact of heavily imbalanced classes of boundary vs. non-boundary regions and the use of distances ($log_{2}$ transformed) as the best representation of the association between TAD boundaries and genomic annotations.

**Methods.** Using replicate Hi-C data for the GM12878 and K562 cell lines, we proposed a novel ensemble framework that addresses these data irregularities and characteristics. Multiple random forest classification models built on regularized sets of genomic annotations, in combination with variations in data resolution, feature engineering, and re-sampling techniques. The composite metric Matthew's Correlation Coefficient (MCC) was used to compare models derived from the framework to determine a consensus approach to predicting TAD boundary formation.

**Results.** We discovered that models built on SMOTE re-sampled data out performed models built simply on  random under- and over-sampled data. Additionally, our results indicated that $log_{2}$ distances between annotations and TAD boundaries were more informative than other predictor-types used to build models. Furthermore, we found that performances tended to degrade as data resolution decreased. Lastly, the most predictive annotations in the best performing models included known architectural proteins such as CTCF, SMC3, RAD21, and ZNF143, as well as lesser known transcription factors including P300 and NFKB.

@@

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed how the spatial organization of the human genome may affect genomic regulation [@lieberman2009comprehensive]. Analyses have shown that the genome is folded into distinct compartments termed topologically associating domains (TADs) [@dixon2012topological]. Approximately 60-80% of TADs remain stable across cell types [@Schmitt:2016aa], suggesting that the boundaries of TADs may play a role in restricting the function of certain genomic elements such as enhancers, thereby impacting the transcription of genes. Furthermore, studies have shown that changing the 3D structure of the DNA, causing disruptions in these domains, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @flavahan2016insulator; @song2017ctcf]. Therefore, it has become increasingly important to be able to identify the key molecular drivers of the formation of TAD boundaries in order to further our understanding of the human genome.

>>>>

**Background.** Chromosome conformation capture sequencing technologies have shown that the three-dimensional (3D) structure of the genome is folded into distinct compartments, known as topologically associated domains (TADs). The locations of TAD boundaries have been shown to be highly conserved. Additionally, the distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify the epigenomic features that are most predictive of TAD boundaries.

Current methods rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions on the linear genome that include TAD boundaries. However, many methods ignore key characteristics of domain data, such as the impact of heavily imbalanced classes distinguished by boundary vs. non-boundary regions and the use of distances ($log_{2}$ transformed) as the best representation of the association between TAD boundaries and genomic annotations.

**Methods.** Using replicate Hi-C data for the GM12878 and K562 cell lines, we proposed a novel ensemble framework that addresses these data irregularities and characteristics. Multiple random forest classification models built on regularized sets of genomic annotations obtained from the UCSC genome browser database, in combination with variations in data resolution, feature engineering, and re-sampling techniques. The composite metric Matthew's Correlation Coefficient (MCC) was used to compare models derived from the framework to determine a consensus approach toward predicting TAD boundary formation.

**Results.** We discovered that models built on SMOTE (Synthetic Minority Over-sampling TEchnique) re-sampled data out performed models built simply on random under- and over-sampled data. Additionally, our results indicated that log2 distances between annotations and TAD boundaries were more informative than other predictor-types used to build models. Importantly, we also found that performances tended to degrade as data resolution decreased. Lastly, the most predictive annotations in the best performing models included known architectural proteins such as CTCF, SMC3, RAD21, and ZNF143, as well as lesser known transcription factors including P300 and NFKB.

@@

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed how the spatial organization of the human genome may affect genomic regulation [@lieberman2009comprehensive]. Analyses have shown that the genome is folded into distinct compartments termed topologically associating domains (TADs) [@dixon2012topological]. Approximately 60-80% of TADs remain stable across cell types [@Schmitt2016aa], suggesting that the boundaries of TADs may play a role in restricting the function of certain genomic elements such as enhancers, thereby impacting the transcription of genes. Furthermore, studies have shown that changing the 3D structure of the DNA, causing disruptions in these domains, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @Flavahan2016; @song2017ctcf]. Therefore, it has become increasingly important to be able to identify the key molecular drivers of the formation of TAD boundaries in order to further our understanding of the human genome.
